Towards Compiler-Agnostic Performance in Finite-Difference Codes

### 2015-09-07 docs/papers/shallow_performance/shallow_perf.tex
Re-word sentence to avoid <IBM results> to please IBM reviewer
PARCO paper: Re-ordering of results section following suggestion of reviewer

The bottom layer, in terms of calling hierarchy, is the Kernel
layer. The Kernel layer implements the science that the Algorithm
layer calls, as a set of subroutines. These kernels operate on local
fields (a set of elements, a single column of elements, or a set of
columns, depending on the kernel). Again the scientist is responsible
for writing this layer and there is no parallelism specified here,
but, depending on the complexity of the Kernels, there may be input
from an HPC expert and/or some coding rules to help ensure the kernels
compile into efficient code.

The PSy layer sits in-between the Algorithm and Kernel layers and its
functional role is to link the algorithm calls to the associated
kernel subroutines. As the Algorithm layer works on logically global
fields and Kernel layer works on local fields, the PSy layer is
responsible for iterating over columns. It is also responsible for
including any required distributed-memory operations, such as halo
swaps and reductions.

@@

For this work we use a benchmark called Shallow which solves the
shallow-water equations on a bi-periodic plane following the
finite-difference scheme introduced by Sadourny~\cite{sadourny75}.
This software was originally written in 1984 by Paul Swarztrauber of
the National Center for Atmospheric Research, US.  However, in common
with many scientific codes, it has subsequently undergone some sort of
evolutionary development with subsequent people making various changes
and optimising it for previous generations of hardware.  In describing
our work, we term the version of the Shallow program obtained at the
beginning of this project the `original' version.

@@

Since any real oceanographic computational model must output results,
we ensure that any \psykal version of Shallow retains the Input/Output
capability of the original. This aids in limiting the optimisations
that can be performed on the \psykal version to those that should also
be applicable to full oceanographic models. Note that although we
retain the I/O functionality, all of the results presented in this work
carefully exclude the effects of I/O since it is compute performance
that interests us here.

@@

The primary aim of this work is to determine whether any performance
loss due to the \psykal separation of concerns can be negated by
optimising the PSy layer. To evaluate this we perform tests on a range
of hardware and a range of compilers. The hardware and compilers are
listed in Table~\ref{TABLE_compilers}. Where a compiler is available
on a given piece of hardware, the version number used in this work is
specified.
%
The Intel Haswell-based system (Xeon E5-1620 v2) has a clock speed of
3.7~GHz and 10~MB of last-level cache. The Intel Ivy Bridge-based
system (Xeon E5-2697) has a clock speed of 2.7~GHz and a last-level
cache of 30~MB. The CPU in the IBM Power-8 system is built around 12
`chiplets'. Each chiplet has a single core with 64~KB of L1 cache,
512~KB of L2 cache and 8~MB of L3 cache. The L3 cache of each chiplet
is shared with all other chiplets so that there is a total of 96~MB of
last-level cache. The cores on the system we used had a clock speed of
3.3~GHz.

@@

Beginning with the vanilla \psykal version, we then apply a series of
(cumulative) code transformations while obeying the \psykal separation
of concerns, {\it i.e.} optimisation is restricted to the middle,
{PS}y layer and leaves the kernel and algorithm layers unchanged. The
aim of these optimisations is to recover, as much as is possible, the
structure and thus performance of the original version of the code. We
began with two transformations that affect all of the kernels. After
this, we determined the transformations to perform by comparing the
timing profiles of the \psykal and original versions of the code. We
then tackled the sections resposible for the largest slowdowns and
iterated this process. While doing this we also made heavy use of the
optimisation reports produced by the compilers and how these changed
when moving from the original to the \psykal version of the code.

@@

@@

Moving now the to the \psykal version of Shallow, it turns out that a
plot of the performance of the best performing \psykal version for
each of the compiler/CPU combinations is very similar to
Figure~\ref{FIG_orig_perf_summary}. Therefore, in
Figure~\ref{FIG_slowdown_summary} we plot the percentage {\em
  difference} between the performance of the original and the (best)
\psykal versions of Shallow for each compiler/CPU combination. The
most important feature of this plot is that the performance penalty
incurred by the \psykal version is less than four percent with only
one exception where it is 6.6\%.  In fact, for the Intel and Cray
compilers, the \psykal version of Shallow is never more than 2\%
slower than the original and in some cases is faster.  This
demonstrates that we can reliably recover the performance of the
original version of the code, despite the significant restructuring
required by the \psykal approach.

Having shown that, in general, we can recover and occasionally improve
upon the performance of the original version of Shallow, the next
logical step is to examine the necessary code transformations.  We do
this for the $256^{2}$ case since this fits within last-level cache on
all of the CPUs we are using here.  The performance of the \psykal
version after each transformation has been applied to the code is
plotted in Figure~\ref{FIG_opt_stages_256}. Results for a given
compiler/CPU are given relative to the performance of the original
version of Shallow with the same compiler/CPU.

@@

@@

The IBM results in particular demonstrate two key points.
%
First, even for a relatively simple
benchmark code, the code structure required to obtain good performance
between different architecture and/or compiler combinations may
differ. For hand optimised codes this implies the need to support
multiple code bases in order to achieve portable performance.
%
Second, the \psykal approach directly addresses the problem of
requiring different optimisations for different architectures and/or
compilers by limiting these differences to a separate performance
layer, thereby allowing the natural science code base to remain
unchanged.

>>>>

The bottom layer, in terms of calling hierarchy, is the Kernel
layer. The Kernel layer implements the science that the Algorithm
layer calls, as a set of subroutines. These kernels operate on fields
that are local to the process doing the computation. (Depending on the
type of kernel, these may be a set of elements, a single column of
elements, or a set of columns.). Again the scientist is responsible
for writing this layer and there is no parallelism specified here,
but, depending on the complexity of the Kernels, there may be input
from an HPC expert and/or some coding rules to help ensure the kernels
compile into efficient code.

The PSy layer sits in-between the Algorithm and Kernel layers and its
functional role is to link the algorithm calls to the associated
kernel subroutines. As the Algorithm layer works on logically global
fields and Kernel layer works on local fields, the PSy layer is
responsible for iterating over columns. It is also responsible for
including any distributed-memory operations resulting from the
decomposition of the simulation domain, such as halo swaps and
reductions.

@@

For this work we use a benchmark called Shallow which solves the
shallow-water equations on a bi-periodic plane following the
finite-difference scheme introduced by Sadourny~\cite{sadourny75}.
This software was originally written in 1984 by Paul Swarztrauber of
the National Center for Atmospheric Research, US.  However, in common
with many scientific codes, it has subsequently undergone some sort of
evolutionary development with subsequent people making various changes
and optimising it for previous generations of hardware.  There is no
complete record of the people involved and the precise hardware they
were targetting. In describing our work, we term the version of the
Shallow program obtained at the beginning of this project the
`original' version.

@@

Since any real oceanographic computational model must output results,
we ensure that any \psykal version of Shallow retains the Input/Output
capability of the original (it reads the problem specification from a
namelist and outputs the solution fields every $n$ timesteps). This
aids in limiting the optimisations that can be performed on the
\psykal version to those that should also be applicable to full
oceanographic models. Note that although we retain the I/O
functionality, all of the results presented in this work carefully
exclude the effects of I/O since it is compute performance that
interests us here.

@@

The primary aim of this work is to determine whether any performance
loss due to the \psykal separation of concerns can be negated by
optimising the PSy layer and if so, what the key transformations are
(for a given system). To evaluate this we perform tests on a range of
hardware and a range of compilers. The hardware and compilers are
listed in Table~\ref{TABLE_compilers}. Where a compiler is available
on a given piece of hardware, the version number used in this work is
specified.
%
The Intel Haswell-based system (Xeon E5-1620 v2) has a clock speed of
3.7~GHz and 10~MB of last-level cache. The Intel Ivy Bridge-based
system (Xeon E5-2697) has a clock speed of 2.7~GHz and a last-level
cache of 30~MB. The CPU in the IBM Power-8 system is built around 12
`chiplets'. Each chiplet has a single core with 64~KB of L1 cache,
512~KB of L2 cache and 8~MB of L3 cache. The L3 cache of each chiplet
is shared with all other chiplets so that there is a total of 96~MB of
last-level cache. The cores on the system we used had a clock speed of
3.3~GHz.

@@

Beginning with the vanilla \psykal version, we then manually apply a
series of (cumulative) code transformations while obeying the \psykal
separation of concerns, {\it i.e.} optimisation is restricted to the
middle, {PS}y layer and leaves the kernel and algorithm layers
unchanged. The aim of these optimisations is to recover, as much as is
possible, the structure and thus performance of the original version
of the code. We began with two transformations that affect all of the
kernels. After this, we determined the transformations to perform by
comparing the timing profiles of the \psykal and original versions of
the code. We then tackled the sections resposible for the largest
slowdowns and iterated this process. While doing this we also made
heavy use of the optimisation reports produced by the compilers and
how these changed when moving from the original to the \psykal version
of the code.

@@

In this section we first discuss the performance of the original
version of the code with the various compiler/CPU combinations. We
then investigate the different code transformations that are required
to recover good performance from the \psykal version of Shallow. Finally,
we compare the most performant \psykal version  with the original
version of Shallow for each compiler/CPU combination.

\subsection{Performance of the Original Version of Shallow}

@@

\subsection{Optimisations of the \psykal Version to Recover Performance}

We now examine the performance of the \psykal version of Shallow and
the code transformations that are required to reach the levels obtained
by the original.  We do this for the $256^{2}$ problem size since this
fits within last-level cache on all of the CPUs we are using here.
The performance of the \psykal version after each transformation has
been applied to the code is plotted in
Figure~\ref{FIG_opt_stages_256}. Results for a given compiler/CPU are
given relative to the performance of the original version of Shallow
with the same compiler/CPU.

@@

\subsection{Performance Comparison of the Original and \psykal Versions}

Finally, we compare the performance of the best \psykal version of
Shallow with that of the original. In
Figure~\ref{FIG_slowdown_summary} we plot the percentage {\em
  difference} between the performance of the original and the (best)
\psykal versions of Shallow for each compiler/CPU combination. The
most important feature of this plot is that the performance penalty
incurred by the \psykal version is less than four percent with only
one exception where it is 6.6\%.  In fact, for the Intel and Cray
compilers, the \psykal version of Shallow is never more than 2\%
slower than the original and in some cases is faster.  This
demonstrates that we can reliably recover the performance of the
original version of the code, despite the significant restructuring
required by the \psykal approach.

@@

In particular, the results on the IBM system demonstrate two key points.
%
First, even for a relatively simple
benchmark code, the code structure required to obtain good performance
between different architecture and/or compiler combinations may
differ. For hand optimised codes this implies the need to support
multiple code bases in order to achieve portable performance.
%
Second, the \psykal approach directly addresses the problem of
requiring different optimisations for different architectures and/or
compilers by limiting these differences to a separate performance
layer, thereby allowing the natural science code base to remain
unchanged.
